---
title: "JT AFC 05/02/2020 - cytofkit / cytofast"
author: "S Granjeaud - A Meghraoui"
date: "05/02/2020"
output:
  html_document:
    toc: true
    fig_height: 9
    fig_width: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project definition

We organize the data and the results

```{r}
fcs_dir = "c:/demo/200205-atelier/CLEAN_DATA"
res_dir = "c:/demo/200205-atelier/CLEAN_DATA_results"
marker_file = "ck_markers_main.txt"
if (!dir.exists(res_dir)) dir.create(res_dir)
```



## FCS Overview


```{r}
library(flowCore)

# read the FCS as a flowset, ie a group of compatible FCS
fs = read.flowSet(path = fcs_dir, pattern = "*.fcs", transformation = FALSE, truncate_max_range = FALSE)

# view the flowset
fs

# view annotations
pData(fs)
# here, none

# view the cell counts
fsApply(fs, nrow)

# view a specific FCS
ff = fs[[1]]  # extract 1st FCS as flowframe
# view some keywords
keyword(ff, "$CYT")  # cytometer
keyword(ff, "$GUID")
keyword(ff, "$FIL")  # file name
keyword(ff, "$TOT")  # cell count
keyword(ff, "$COM")  # comment
# view all keywords
kwd = keyword(ff)
length(kwd)  # there are too many
# filter out some keywords
kwd_1 = kwd[!grepl("flowCore", names(kwd))]  # not flowcore
kwd_2 = kwd_1[-grep("^\\$P", names(kwd_1))]  # not standard parameters
#kwd_2  # display remaining keywords
unlist(kwd_2)
```


## cytofkit analysis


Run by hand

```{r eval=FALSE}
library(cytofkit)
# Launch the Graphical User Interface for tuning the run
cytofkit_GUI()
# Note the path to the result file

# Launch the Shiny interface to view and annotate the analysis
cytofkitShinyAPP()
```


```{r eval=FALSE}
# Launch the Shiny interface using a defined path
analysis_file = "c:/demo/200205-atelier/CLEAN_DATA_results/run_5k/run_5k.RData"
if (file.exists(analysis_file))
  cytofkitShinyAPP(analysis_file)
```


Script version for tuning parameters

```{r eval=FALSE, include=FALSE}
library(cytofkit)
if (!file.exists(file.path(fcs_dir, marker_file)))
  stop("Cannot find marker file!")
cytofkit(fcsFiles = fcs_dir, markers = marker_file, projectName = "run_5k", ifCompensation = FALSE, transformMethod = "arcsinh", mergeMethod = "fixed", fixedNum = 5000, dimReductionMethod = "tsne", clusterMethods =  c("FlowSOM", "Rphenograph"), visualizationMethods = c("pca", "tsne", "umap"), resultDir = file.path(res_dir, "run_5k"))
analysis_file = file.path(res_dir, "run_5k", "run_5k.RData")
if (file.exists(analysis_file))
  cytofkitShinyAPP(analysis_file)
```


## Cytofast post-analysis


Main configuration

```{r}
# Tell where the FCS files of cytofkit are
analysis_fcs_dir = file.path(res_dir, "run_5k", "run_5k_analyzedFCS")
stopifnot(dir.exists(analysis_fcs_dir))
selected_clustering_method = "FlowSOM_clusterIDs"
#selected_clustering_method = "Rphenograph_clusterIDs"
```

Read Cytofkit FCS export

```{r readCytofkit}
# Load libraries
library(flowCore)
library(cytofast)

# Load cytofkit FCS files
cfData = readCytofkitFCS(analysis_fcs_dir, clusterID = selected_clustering_method)
# cfData = readCytofkitFCS("demo/cytofkit_demo_analyzedFCS", clusterID = "Rphenograph_clusterIDs")
table(cfData@expr$clusterID)
```

Optionnally, do some channel cleaning

```{r}
# Remove unneeded channels
keep <- colnames(cfData@expr)  # all
keep <- keep[!grepl("^(Time|Event_length|Cell_length)", keep, ignore.case = TRUE)]
keep <- keep[!grepl("^(Cisplatin)", keep, ignore.case = TRUE)]
keep <- keep[!grepl("(_ADN1|_ADN2)", keep, ignore.case = TRUE)]
keep <- keep[!grepl("^(NA\\.)", keep, ignore.case = TRUE)]
keep <- keep[!grepl("^X\\d+", keep) | grepl("_", keep)]  # unannotated X channel
cfData@expr <- cfData@expr[, keep]

# Rename channels
colnames(cfData@expr) <- gsub("(^X\\d+.+?_)", "", keep)  # remove metal tag
colnames(cfData@expr) <- gsub("(\\.\\d+.$)", "", keep)  # remove metal tag

# Manual ordering of channels
first <- c("CD20", "CD19", "CD3", "CD4", "CD8", "TCRgd", "CD56", "CD16", "CD45RA", "CD95", "CD127")
first <- intersect(first, colnames(cfData@expr))

keep <- colnames(cfData@expr)  # all
cfData@expr <- cfData@expr[, c(keep[1:2], first, setdiff(keep[-(1:2)], first))]
```


Read meta data that group FCS samples

```{r}
# To compute statistical tests, we need to define groups
# Here are the un-annotated samples
cfData@samples

# Either we fill the annotation using an Excel template
# the template consists in a first column of sample identifiers
# and a second column called status
meta <- data.frame(cfData@samples, status = "")
# The template is written on disk
if (!file.exists("meta.csv")) {
  write.csv(meta, "meta.csv")
}
# Now you can annotate the template using Excel
# and store it as CSV file keeping the original format

# Either we fill meta data programmatically
meta$status <- c(rep(c("D", "P"), each = 3))
write.csv(meta, "meta.csv")

# Fill the template using Excel or so

# Import annotation
if (file.exists("meta.csv")) {
  meta <- read.csv("meta.csv", row.names = 1)
}
# meta <- meta[match(row.names(cfData@samples), meta[,"sampleID"]),] # match sampleID
cfData@samples <- cbind.data.frame(cfData@samples, meta[, -1, drop = FALSE])
cfData@samples
```

Transform expression values, in the same way you display marker intensity

```{r}
# Transform expression with asinh( x / 5 )
cfData@expr[,-(1:2)] <- asinh(cfData@expr[,-(1:2)]/5)
```

Works with cell counts

```{r}
# Retrieve cell counts
cfData <- cellCounts(cfData)
head(cfData@counts)
# View(cfData@counts)
cytoHeatmaps(cfData, group="status", legend=TRUE)
```

Cell counts need to be transformed

```{r}
# Store raw counts
cellCountRaw <- cfData@counts

# Transform cell counts with log2( x + floor )
cfData@counts <- log2(cfData@counts+10)
# Center
cfData@counts <- sweep(cfData@counts, 2, colMeans(cfData@counts), "-")
round(head(cfData@counts))
cytoHeatmaps(cfData, group="status", legend=TRUE)
```

Let's take status D as reference

```{r}
# Center to D
cfData@counts <- sweep(cfData@counts, 2, colMeans(cfData@counts[cfData@samples$status == "D",]), "-")
round(head(cfData@counts)*10)
cytoHeatmaps(cfData, group="status", legend=TRUE)
```

Box plots

```{r}
# Detailed view of counts aka percentages
cytoBoxplots(cfData, group = "status")
```

Functional marker histograms

```{r, eval=FALSE}
# Detailed of functional markers
msiPlot(cfData, markers = c("CD8", "CD4"), byGroup='status')
```

Do some statistics

```{r}
# Add t.test
cfData <- cytottest(cfData, group="status", adjustMethod = "fdr")
head(cfData@results)

# Add some columns
# fold is a multiplicative coefficient applied to the reference
# fold is negative to stand for 1/fold
cfData@results <- within(cfData@results, {
  diff <- mean_P - mean_D
  fold <- sign(diff) * round(2^abs(diff), 2)
  unselected <- abs(diff) < 0.5 | pvalue > 0.05
  label <- clusters
  label[unselected] <- NA
  })
```

View the Volcano plot

```{r}
library(ggplot2)
library(ggrepel)
set.seed(42)

p <- ggplot(cfData@results, aes(diff, -log10(pvalue))) + geom_point() + geom_hline(yintercept = -log10(c(0.05, 1))) + geom_vline(xintercept = c(-0.5, 0 , 0.5)) + ggtitle("Volcano Plot, diff = Log2 Fold Change of percentages")
p + geom_text_repel(aes(label = label))

```

Alternate scale

```{r}
p <- ggplot(cfData@results, aes(2^diff, -log10(pvalue))) + geom_point() + geom_hline(yintercept = -log10(c(0.05, 1))) + geom_vline(xintercept = c(1/1.5, 1 , 1.5)) + ggtitle("Volcano Plot, diff = Log2 Fold Change of percentages") + scale_x_continuous(trans = "log2")
p + geom_text_repel(aes(label = label))
```


and the table of counts

```{r}
library(knitr)
kable(
  cbind(cfData@results[with(cfData@results, !unselected),], t(
    cellCountRaw[,with(cfData@results, !unselected)]))[,-c(1:3,6:7)]
)
```




